Learning to Synthesize Rules for Network Intrusion Detectors

### 2020-02-23 paper/paper.tex
pass

\section{Illustrative Examples}
\label{sec:suri-metas-coverage}

This section briefly presents examples of attacks and illustrates the
rules that \tname{} synthesizes to capture those attacks.

\subsection{Active Reconnaissance}
\label{sec:active-recon}

%% \Mar{@Lucas, acima falamos de porta. Abaixo nao falamos de
%%   porta, apenas de maquina. Vc. pode ser mais especifico? ->} \Luc{De
%%   fato estamos falando de duas coisas diferentes, mas ambas são
%%   exemplos de 'active reconnaissance'. Primeiramente citamos o port
%%   scanning, q tem o objetivo de encontrar portas abertas em uma
%%   maquina especifica. Depois falamos do ping scan, q tem o objetivo de
%%   encontrar ip's ativos numa rede.}

Active Reconnaissance is a method used by malicious individuals to
collect information about a computing system to determine potential
vulnerabilities. This methods is used as a preliminary step towards an
actual attack. For example, Ping Scan scans the network for IPs of
accessible hosts. An attacker sends several ICMP Echo requests to a
range of IP addresses in the network and checks which ones respond to
these requests. This method enables an attacker to know which machines
are active in the network.

@@

Figure 4 shows a Suricata rule that captures a Ping Scan attempt, as
documented in NMAP~\cite{netmap}, a network discovery and security
auditing open-source tool. The relevant options are in bold---the
option \CodeIn{dsize: 0} indicates that the packet has no bytes in the
payload and the option \CodeIn{itype: 8} indicates that the packet to
be captured are ICMP Echo requests.

In the following, we discuss the rule produced by \tname{} to capture
this attack. The action and the header of the rules that \tname{}
creates are either inferred or left constant. For example, in this
case, \tname{} sets the action to \CodeIn{alert} and sets the header
to \CodeIn{<proto> any any -> any any}, indicating that it infers the
target protocol and uses the most general flow description possible,
\ie{}, it instructs the \nids\ to analyze the traffic flowing from/to
any address and port.  The choice of these parameters are
domain-specific and subject to change by the system administrator.
The main challenge in synthesizing rules is to discover the set of
options in a rule that blocks the malicious traffic, but allows the
benign traffic to pass.

\tname{} proceeds as follows to synthesize rules. First, it analyzes the malicious
traffic and extracts options expressed in that traffic. Recall from
Section~\ref{sec:rules-and-packets} that rule options can be inferred
from packet contents. \tname{} finds the following set of options at
this stage:

\begin{figure}[h]
  \vspace{-2ex}
  \lstinputlisting[language=C,numbers=none,frame=none,keywords={dsize,icode,itype,icmp\_id,icmp\_seq}]{pingscan.suricata.synth}
  \vspace{-2ex}  
\end{figure}

\noindent
By construction, a rule that uses this set of options captures the
malicious traffic. Unfortunately, the rule above is
overspecified. Informally, the rule includes ``too specific''
characteristics of the malicious traffic and could potentially result
in false negatives. In this case, although options \CodeIn{dsize:0}
and \CodeIn{itype:8}~(see Figure~\ref{fig:pingscan-example}) are
present in the set of inferred options, the others are not.

Second, \tname{} uses the benign traffic to mitigate the overfitting
issue. In this step, \tname{} searches for alternative rules that
preserve the invariant to capture the positive malicious traffic but
not capture the negative benign traffic. It uses the rule obtained in
the previous step as seed to bootstrap the search for alternative
plausible rules, \ie{}, rules that satisfy the aforementioned
invariant. The search is designed to generate rules containing subsets
of the options from the seed rule. Considering this example, \tname{}
produces a total of \Fix{XXXXX} plausible rules at the end of the
search.

Finally, \tname{} ranks plausible rules according to heuristics
learned from a database of existing rules. The intuition is that new
rules should share characteristics with pre-existing rules. For
example, one heuristic function used by \tname{} leverages the amount
of options a rule typically has. The score of a rule, given by this
function, is proportional to the popularity of the number of options
declared in rules of a given protocol. At the end of the process,
\tname{} ranks the golden rule at the top of the ranking including
\Fix{XXXXXX} other plausible rules.

%% the following rules 
%% at the top of the ranking produced by \tname{}:
%% 1)~{\small\CodeIn{itype:8; dsize:0;}}, 2)~{\small\CodeIn{itype:8;
%%     icmp\_id:23570;}}, 3)~{\small\CodeIn{icode:8; icmp\_id:23570;}},
%% 4)~{\small\CodeIn{itype:8; icode:0; dsize:0;}},
%% 5)~{\small\CodeIn{icmp\_id:23570}}. Note that the golden rule appears
%% at the top position.

%% \begin{figure}[h]
%%   \vspace{-2ex}
%%   \lstinputlisting[language=C,numbers=none,frame=none,keywords={dsize,icode,itype,icmp\_id,icmp\_seq}]{top-pingscan.rules}
%%   \vspace{-3ex}  
%% \end{figure}

To sum up, \tname{} uses the messages that manifest an attack in the
first step to create a potentially overspecified seed rule, it uses
the benign traffic in the second step to create a set of plausible
rules from the seed rule, and it uses public databases with existing
rules in the third step to rank plausible rules.

\subsection{Remote Code Execution (RCE)}
\label{sec:rce}
\label{sec:jboss}
\label{sec:content-example}

\subsection{Denial-of-Service}
\label{sec:dos}

@@

>>>>

\section{Illustrative Example}
\label{sec:suri-metas-coverage}
\label{sec:active-recon}

This section briefly shows how \tname{} synthesizes rules for one
example attack. Active Reconnaissance is a method used by malicious
individuals to collect information about a computing system to
determine potential vulnerabilities. This method is used as a
preliminary step towards an actual attack. For example, Ping Scan
scans the network for IPs of accessible hosts. An attacker sends
several ICMP Echo requests to a range of IP addresses in the network
and checks which ones respond to these requests. This method enables
an attacker to know which machines are active in the network.

@@

Figure~\ref{fig:pingscan-example} shows a Suricata rule that captures
a Ping Scan attempt, as documented in NMAP~\cite{netmap}, a network
discovery and security auditing open-source tool. The relevant options
(see Section~\ref{sec:example-suricata-rules}) are in bold---the
option \CodeIn{dsize: 0} indicates that the packet has no bytes in the
payload and the option \CodeIn{itype: 8} indicates that the packet to
be captured are ICMP Echo requests.

In the following, we briefly show how \tname{} synthesizes rules for
this attack. The action and the header of the rules that \tname{}
creates are either inferred or left constant. For example, in this
case, \tname{} sets the action to \CodeIn{alert} and sets the header
to \CodeIn{<proto> any any -> any any}, indicating that it infers the
target protocol and uses the most general flow description possible,
\ie{}, it instructs the \nids\ to analyze the traffic flowing from/to
any address and port.  The rationale for this decision is that the
choice of these parameters are domain-specific and subject to change
by the system administrator.  The main challenge in synthesizing rules
is to discover the set of options in a rule that blocks the malicious
traffic, but allows the benign traffic to pass.

\tname{} proceeds as follows to synthesize rules. First, it analyzes
the malicious traffic and extracts options expressed in that
traffic. Recall from Section~\ref{sec:rules-and-packets} that some
options can be inferred from packet contents. \tname{} finds the
following options at this stage:
{\scriptsize{\texttt{\textbf{dsize}:0; \textbf{itype:}8;
      \textbf{icode:}0; \textbf{icmp\_id:}23570;
      \textbf{icmp\_seq:}3439;}}}.  By construction, a rule that uses
this set of options captures the malicious traffic. Unfortunately, the
rule above is overspecified. Informally, the rule includes ``too
specific'' characteristics of the malicious traffic and could
potentially result in false negatives. In this case, although the
options \CodeIn{dsize:0} and \CodeIn{itype:8} are present in the set
of inferred options (as the golden rule from
Figure~\ref{fig:pingscan-example}), the others are not.

%% \vspace{-2.5ex}
%% \begin{figure}[h]
%%   \lstinputlisting[language=C,numbers=none,frame=none,keywords={dsize,icode,itype,icmp\_id,icmp\_seq}]{pingscan.suricata.synth}
%% \end{figure}
%% \vspace{-2.5ex}
%% \noindent

Second, \tname{} uses the benign traffic to mitigate the overfitting
issue. In this step, \tname{} searches for alternative rules that
preserve the invariant to capture the positive malicious traffic but
not capture the negative benign traffic. It uses the rule obtained in
the previous step as seed to bootstrap the search for alternative
plausible rules, \ie{}, rules that satisfy the aforementioned
invariant. The search is designed to generate rules containing subsets
of the options from the seed rule. Considering this example, \tname{}
produces a total of \Fix{XXXXX} plausible rules at the end of the
search.\Mar{Technically, there are $32 (=2^5)$ subsets for a set with 5
  elements. So, it is important to indicate why Table 3 shows sooo
  many plausible rules and that this is is a very simple rule (for illustration).}

Finally, \tname{} ranks plausible rules according to heuristics
learned from a database of existing rules. The intuition is that new
rules should share characteristics with pre-existing rules. For
example, one heuristic function used by \tname{} leverages the amount
of options a rule typically has. The score of a rule, given by this
function, is proportional to the popularity of the number of options
declared in rules of a given
protocol. Figure~\ref{fig:distribution-contents} shows the
distribution of the number of content options for the database of
rules we used\Mar{reference}. Note that rules with 1 to 3 options are
more common.  At the end of the process, \tname{} ranks the golden
rule at the top of the ranking including \Fix{XXXXXX} other plausible
rules.

\Mar{add short par indicating that other problems can be much more
  complex than this...}

To sum up, \tname{} uses the messages that manifest an attack in the
first step to create a potentially overspecified seed rule. Then, it
uses benign traffic in the second step to create a set of plausible
rules from the seed rule. Finally, in the third step, \tname{} uses
existing rules to learn heuristics to rank plausible rules.

@@

\section{Discussion}

\subsection{Sample of other attacks}

\subsubsection{Remote Code Execution (RCE)}
\label{sec:rce}
\label{sec:jboss}
\label{sec:content-example}

\subsubsection{Denial-of-Service}
\label{sec:dos}
